Water discharge has a variable impact on sediment transport capacity in subglacial systems

### 2023-07-13 Delaney2023.tex
more bits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
% 
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
% 
% You may copy this file and give it your
% article name, and enter your text.
% 
% 
% Step 1: Set the \documentclass
% 
%

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage{hyphenat}
\usepackage{float}
\usepackage{multirow}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\arraybackslash\hspace{0pt}}p{#1}}

\linenumbers
\sloppy
\hbadness=99999 
%%%%%%% 
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
% 
% \note[editor]{The note}
% \annote[editor]{Text to annotate}{The note}
% \add[editor]{Text to add}
% \remove[editor]{Text to remove}
% \change[editor]{Text to remove}{Text to add}
% 
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

\draftfalse

%% Enter journal name below.
%% Choose from this list of Journals:
% 
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
% 
% ie, \journalname{Water Resources Research}

\journalname{Geophysical Research Letters}

@@

\newenvironment{laysummary}{
  \chapter{\centering \textbf{Plain Language Summary}}
}

@@

\title{Water discharge has a variable impact on sediment transport capacity in subglacial systems}

%% ------------------------------------------------------------------------ %%
% 
% AUTHORS AND AFFILIATIONS
%

% Authors are individuals who have significantly contributed to the
% research and preparation of the article. Group authors are allowed, if
% each author in the group is separately identified in an appendix.)

% List authors by first name or initial followed by last name and
% separated by commas. Use \affil{} to number affiliations, and
% \thanks{} for author notes.
% Additional author notes should be indicated with \thanks{} (for
% example, for current addresses).

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antarctica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}

\authors{Ian Delaney\affil{1}}

\affiliation{1}{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, CH-1015 Lausanne }

% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

% \affiliation{=number=}{=Affiliation Address=}
% (repeat as many times as is necessary)

%% Corresponding Author:
% Corresponding author mailing address and e-mail address:

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)

% Example: \correspondingauthor{First and Last Name}{email@address.edu}

\correspondingauthor{Ian Delaney}{ianarburua.delaney@unil.ch}

@@

\begin{keypoints}
\item Water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size, given their slow evolution.

@@

\item The systems' divergence in transport response may impact sediment dynamics at glacier margins and interpretation of sediment records.
\end{keypoints}

@@

%%% Suggested section heads:
\ian{comparison with Alley 1997... mention his variability-- also deal with his $Qs \propto Qw^{\frac{9}{2}}$  treats like rivers?  does not consider channel size evolution}

@@

Glaciers expel large quantities of sediment from their termini \cite{hallet1996}.
Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>[]{zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some of these regions, increased water discharge and glacier melt have been attributed to greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
In other cases, enhanced access of meltwater to subglacial sediment at high elevations was attributed as the cause of increased sediment transport \cite{delaney2020,vergara2022}.
Observed changes to sediment transport in glacierized catchments create an imperative to examine the processes controlling sediment discharge here. 
Yet, the relationship between water discharge and sediment transport in subglacial environments remains minimally discussed and not contrasted with the subaerial river systems that these glaciers feed.

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from below glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}. 
On shorter time periods, pressurized water transports sediment from the below glacier \cite{walder1994,creyts2013,beaud2018}, should enough sediment be present below the glacier (i.e. in a transport-limited regime).

In a transport-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \cite{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
% 
where $w_c$ is the width of the channel.
Following the conservation of mass, the velocity of the water flowing through a  channel is given as 
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q_w}{S},
  \end{equation}
\end{linenomath*}
% 
where $Q_w$ is water discharge,  and $S$ is the channel's cross-sectional area.

In subaerial channels, with open channel flow, the cross-sectional area $S$ of the channel evolves with changing water discharge $Q_w$  by changing the width of the channel along with the depth \cite{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, the cross-sectional area $S$ increases as the channel widens and deepens, along with water velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both the channel width $w_c$  and water velocity $v$, which is squared.

Yet, the response of water velocity to changing water discharge in subglacial channels is different than subaerial ones.
In subglacial systems, the water is pressurized by the ice above \cite{rothlisberger1972}.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size evolves over days to months, whereas water discharge can vary over hours \cite<e.g.>[]{iken1986,andrews2014,nanni2020}.
In turn, changes in water discharge $Q_w$ are mainly accommodated with water velocity $v$, because the size of the channel $S$ changes on a much shorter timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q_w$ will cause the shear stress across the channel $\tau_t$ to respond to the water velocity $v$, as opposed to the subaerial system where channel width $w_c$ evolves as well.

As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial systems  \cite<e.g.>[]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
However, some suggest that subglacial sediment discharge  $Q_s \propto Q_w^{\frac{9}{2}}$, assuming that water discharge, changing channel width and depth are each proportionally responsible for the variance of water velocity and sediment transport capacity \cite{alley1997,swift2021}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>[]{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \cite<e.g.>[]{muller1968,richards2003,ganti2016}.
Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.
Furthermore, thorough observations of water velocity and channel width in subglacial systems are limited and not contrasted with subaerial conditions at the glacier margins.

To better understand the relationship between sediment transport capacity and water discharge in subglacial systems, this manuscript leverages parameterizations of subglacial and subaerial hydraulics.
These parameterizations are applied to hydrological records from the Fieschergletscher, an alpine glacier in Switzerland, and Leverett glacier,  a land-terminating glacier in Greenland.
Outputs are used to explore differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial. 
With these parameterizations, the manuscript examines the impact of water discharge variability on sediment transport capacity in these subglacial and subaerial systems.
The manuscript then uses the demonstrated differences between the two systems to discuss the implications for sediment transport processes in subglacial systems, compared to subaerial ones, and for interpreting records of sediment transport from glacierized catchments.

\begin{center}
  \begin{figure}[H]
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$, and subaerial, $v_f$, systems are shown by arrow length. $S_g$ and $S_f$ represent the cross-sectional area in subglacial and subaerial channels. The subglacial channel widths $w_{cg}$ remain steady, while the subaerial channel widths evolve $w_{cf}$, denoted by the thick black line. Shear stress $\tau$ is responsible for the mobilization of sediment and increases with the number of makers at the channel bed. 
      Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
    \label{fig:cartoon}
  \end{figure}
\end{center}

\section{Methods}
\label{sect:meth}
Parameterizations below represent relationships amongst water discharge, subglacial water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Because sediment transport largely depends on shear stress applied by water flow \cite{shields1936}, both parameterizations calculate this value and integrate  it across the channel bed (Figure \ref{fig:cartoon}).
The evaluation of shear stress omits a grain size parameter and, thus, leaves open the selection of a sediment transport relationship  \cite<e.g.>[]{shields1936,meyer1948} to  calculate sediment transport capacity.
As a result, relative relationships in shear stress, as opposed to sediment transport capacity, are examined.

\ian{update hooke angle }
\begin{table}[H]
  \centering
  \caption{Variables, parameters, and constants }
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\ \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q_w$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
    Channel cross-sectional area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Shear stress (glacier, subaerial) & $\tau_g,\,\tau_f$&& $\mathrm{Pa \, m^{-1}}$ \\
    Width-integrated shear-stress (glacier, subaerial) & $\tau_{tg},\, \tau_{tf}$&& $\mathrm{Pa \, m^{-1}}$ \\

@@

\end{table}
  $^a$ for Fieschergletscher

  $^b$ for Leverett glacier

@@

To evaluate the shear stress of water flowing across sediments below a glacier, the subglacial channel parameterization evaluates the channel geometry and the velocity of the flowing water.
To accomplish this, a modified version of  the lumped hydraulics model presented in \citeA{clarke1996} and \citeA{werder2010} is used.

Here, it is assumed that the water is transported through a subglacial channel flowing down the glacier; \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to frictional heating from water flow and creep closure by ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>[]{rothlisberger1972,clarke2003}.
In the formulation here does not consider englacial storage of water, and thus that changes in hydraulic head at the top of the glacier are negligable in the evolution of the channel size. 
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_2 \frac{Q \Delta h}{l} - C_2 (h_{o}-\frac{\Delta h}{2})^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure, and $n$ is Glen's n.
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the second term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach, head drop $\Delta h$ is,
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{g}\,s\,f_r\,\frac{Q^2}{D_h^5}
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, and $l$ is the length overwhich head change occurs. $s$ is a factor accounting for channel geometry \cite{hooke1990}, calculated as:
\begin{equation}
  \label{eq:Hf}
  s = \frac{2\,(\beta -\sin \beta)^2}{(\frac{\beta}{2}\,+\,\sin \frac{\beta}{2})^4},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle). Note that $\beta =\pi$ corresponds to a semi-circle and
smaller values of $\beta$ result in shallow, wide channels.

@@

The width of the flat channel floor $w_{cg}$, given angle $beta$ is 
\begin{equation}
  \label{eq:dh2wc}
  w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S}{\beta -\sin \beta}}.
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Here, it is assumed that there is a conservation of mass and a sufficiently wide channel so that the hydraulic radius is consistent with flow depth, uniform flow, and the Darcy-Weisbach relationship.
The shear stress $\tau_f$ at the river bed is represented as
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q_w}{w_{cf}} \Big)^{\frac{2}{3}} \,\frac{\partial z_c}{\partial x}^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\frac{\partial z_c}{\partial x}$ is the channel slope, and $f_p$ is the friction factor for subaerial systems.
Channel width $w_{cf}$ is 
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q_w^e,
\end{equation}
% 
where $k$ is a constant and $e$ is an exponent commonly equal to $\frac{1}{2}$ \cite{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
% 
As in Section~\ref{sect:sub_mode}, the width integrated shear stress is $\tau_{tf}=w_{cf}\,\tau_f$.

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher in the Swiss Alps (info) and the Leverett glacier on the Greenland Ice Sheet (info).
The hydrographs applied to the subglacial and subaerial parameterizations represent the subglacial flow of water as it leaves the glacier and moves into a subaerial flow in front of the glacier, as water moves through the catchment.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, the Fieschergletscher case is referred to as \alpine{} and is exemplified by relatively thin ice thickness ($h_o$: xxx), low discharge (< $20$\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in discharge.
The Leverett glacier case is referred to as \icesheet{} and is exemplified by thick ice  ($h_o$: xxx), high water discharge (< $300$\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in discharge.

The parameterizations are first applied to a reference test case for each glacier that assumes a semi-circular subglacial channel ($\beta=\pi$) and friction factors $f_r$ and $f_p$ are tuned so that the model reproduces reasonable velocities for both  the Greenland Ice Sheet and alpine glaciers \cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
Water velocity, shear-stress and width-integrated shear-stress in the subglacial and subaerial systems are presented below.

>>>>

\documentclass[11pt]{article}
\usepackage [colon,square]{natbib}
\usepackage {comment}
\usepackage{cite}
\usepackage[mathlines]{lineno} %% <- mathlines turns on line numbering in equations
\usepackage{amsmath} 
\usepackage{printlen}
\setlength{\parindent}{10pt}
\setlength{\parskip}{5pt}
\usepackage[explicit]{titlesec}
\usepackage{helvet}
\usepackage{array}
\usepackage{printlen}
\usepackage[a4paper, margin=1.25cm]{geometry}
\usepackage{color,hyperref}
\catcode`\_=11\relax
\newcommand\email[1]{\_email #1\q_nil}
\def\_email#1@#2\q_nil{%
  \href{mailto:#1@#2}{{\emailfont #1\emailampersat #2}}
}
\newcommand\emailfont{\sffamily}
\newcommand\emailampersat{{\color{red}\small@}}
\catcode`\_=8\relax

\usepackage{hyperref}
\usepackage[export]{adjustbox}
\usepackage{bibentry}
\renewcommand{\bibfont}{\footnotesize}
\usepackage{tikz}
\usepackage{setspace}
\usepackage[inline,shortlabels]{enumitem}
\usepackage[all]{nowidow}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
\pagenumbering{gobble} 
\hypersetup{ % Remove ugly boxes around URLs
  colorlinks, linkcolor={black!100!black}, citecolor={black!50!black}, urlcolor={black!80!black}
}

\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\arraybackslash\hspace{0pt}}p{#1}}

\linenumbers
\sloppy
\hbadness=99999

@@

@@

\begin{center}
\Large{\textbf{Water discharge quantity has a variable impact on sediment transport capacity in subglacial systems}}
\normalsize

Ian Delaney

Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, CH-1015 Lausanne

ianarburua.delaney@unil.ch

\end{center}

@@

\paragraph{Keypoints}
\begin{itemize}
\item Water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size, given their slow evolution.

@@

\item The systems' divergence in transport response may impact sediment dynamics at glacier margins and interpretation of sediment records.
\end{itemize}

@@

%%% Suggested section heads:
%TODO: 
%Check with Daniel and Mauro about data

Recheck results figure names.

@@

Glaciers expel large quantities of sediment from their termini \citep{hallet1996}.
Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
Increases in sediment transport  have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
In some of these regions, increased water discharge and glacier melt have been attributed to greater sediment transport capacity \citep{bendixen2017,costa2017,li2021}.
In other cases, enhanced access of meltwater to subglacial sediment at high elevations was attributed as the cause of increased sediment transport \citep{delaney2020,vergara2022}.
Observed changes to sediment transport in glacierized catchments create an imperative to examine the processes controlling sediment discharge here. 
Yet, the relationship between water discharge and sediment transport in subglacial environments remains minimally discussed and not contrasted with the subaerial river systems that these glaciers feed.

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from below glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}. 
On shorter time periods, pressurized water transports sediment from the below glacier \citep{walder1994,creyts2013,beaud2018}, should enough sediment be present below the glacier (i.e. in a transport-limited regime).

In a transport-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \citep{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
% 
where $w_c$ is the width of the channel.
Following the conservation of mass, the velocity of the water flowing through a  channel is given as 
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q_w}{S},
  \end{equation}
\end{linenomath*}
% 
where $Q_w$ is water discharge,  and $S$ is the channel's cross-sectional area.

In subaerial channels, with open channel flow, the cross-sectional area $S$ of the channel evolves with changing water discharge $Q_w$  by changing the width of the channel along with the depth \citep{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, the cross-sectional area $S$ increases as the channel widens and deepens, along with water velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both the channel width $w_c$  and water velocity $v$, which is squared.

Yet, the response of water velocity to changing water discharge in subglacial channels is different than subaerial ones.
In subglacial systems, the water is pressurized by the ice above \citep{rothlisberger1972}.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \citep{rothlisberger1972}.
As a result, the subglacial channel size evolves over days to months, whereas water discharge can vary over hours \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
In turn, changes in water discharge $Q_w$ are mainly accommodated with water velocity $v$, because the size of the channel $S$ changes on a much shorter timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q_w$ will cause the shear stress across the channel $\tau_t$ to respond to the water velocity $v$, as opposed to the subaerial system where channel width $w_c$ evolves as well.

As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial systems  \citep[e.g.][]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
However, some suggest that subglacial sediment discharge  $Q_s \propto Q_w^{\frac{9}{2}}$, assuming that water discharge, changing channel width and depth are each proportionally responsible for the variance of water velocity and sediment transport capacity \citep{alley1997,swift2021}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \citep[e.g.][]{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \citep[e.g.][]{muller1968,richards2003,ganti2016}.
Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.
Furthermore, thorough observations of water velocity and channel width in subglacial systems are limited and not contrasted with subaerial conditions at the glacier margins.

To better understand the relationship between sediment transport capacity and water discharge in subglacial systems, this manuscript leverages parameterizations of subglacial and subaerial hydraulics.
These parameterizations are applied to hydrological records from an alpine glacier in Switzerland (Fieschergletscher), and  a land-terminating glacier in Greenland (Leverett glacier).
Outputs are used to explore differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial. 
With these parameterizations, the manuscript examines the impact of water discharge variability on sediment transport capacity in these subglacial and subaerial systems.
The manuscript then uses the demonstrated differences between the two systems to discuss the implications for sediment transport processes in subglacial systems, compared to subaerial ones, and for interpreting records of sediment transport from glacierized catchments.

\begin{center}
  \begin{figure}[h]
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$, and subaerial, $v_f$, systems are shown by arrow length. $S_g$ and $S_f$ represent the cross-sectional area in subglacial and subaerial channels. The subglacial channel widths $w_{cg}$ remain steady, while the subaerial channel widths evolve $w_{cf}$, denoted by the thick black line. Shear stress $\tau$ is responsible for the mobilization of sediment and increases with the number of makers at the channel bed. 
      Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
    \label{fig:cartoon}
  \end{figure}
\end{center}

\section{Methods}
\label{sect:meth}
Parameterizations below represent relationships amongst water discharge, subglacial water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Because sediment transport largely depends on shear stress applied by water flow \citep{shields1936}, both parameterizations calculate this value and integrate  it across the channel bed (Figure \ref{fig:cartoon}).
The evaluation of shear stress omits a grain size parameter and, thus, leaves open the selection of a sediment transport relationship  \citep[e.g.][]{shields1936,meyer1948} to  calculate sediment transport capacity.
As a result, relative relationships in shear stress, as opposed to sediment transport capacity, are examined.

\begin{table}[ht]
  \centering
  \caption{Variables, parameters, and constants }
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\ \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q_w$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
    Channel cross-sectional area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Shear stress (glacier, subaerial) & $\tau_g,\,\tau_f$&& $\mathrm{Pa \, m^{-1}}$ \\
    Width-integrated shear-stress (glacier, subaerial) & $\tau_{tg},\, \tau_{tf}$&& $\mathrm{Pa \, m^{-1}}$ \\

@@

  $^a$ for \alpine{} (Fieschergletscher)

  $^b$ for \icesheet{} (Leverett glacier)

\end{table}

@@

To evaluate the shear stress of water flowing across sediments below a glacier, the subglacial channel parameterization evaluates the channel geometry and the velocity of the flowing water.
To accomplish this, a modified version of  the lumped hydraulics model presented in \citet{clarke1996} and \citet{werder2010} is used.

Here, it is assumed that the water is transported through a subglacial channel flowing down the glacier; \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972}, and that the channel  size responds to frictional heating from water flow and creep closure by ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \citep[e.g.][]{rothlisberger1972,clarke2003}.
The formulation here does not consider englacial storage of water, and thus that changes in hydraulic head at the top of the glacier are negligable in the evolution of the channel size. 
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_2 \frac{Q \Delta h}{l} - C_2 (h_{o}-\frac{\Delta h}{2})^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure, and $n$ is Glen's n.
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the second term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach, head drop $\Delta h$ is,
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{g}\,s\,f_r\,\frac{Q^2}{D_h^5}
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, and $l$ is the length overwhich head change occurs. $s$ is a factor accounting for channel geometry \citep{hooke1990}, calculated as:
\begin{equation}
  \label{eq:Hf}
  s = \frac{2\,(\beta -\sin \beta)^2}{(\frac{\beta}{2}\,+\,\sin \frac{\beta}{2})^4},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle). Note that $\beta =\pi$ corresponds to a semi-circle and
smaller values of $\beta$ result in shallow, wide channels.

@@

The width of the flat channel floor $w_{cg}$, given angle $\beta$ is 
\begin{equation}
  \label{eq:dh2wc}
  w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S}{\beta -\sin \beta}}.
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citet{tucker1997} is implemented.
Here, it is assumed that there is a conservation of mass and a sufficiently wide channel so that the hydraulic radius is consistent with flow depth, uniform flow, and the Darcy-Weisbach relationship.
The shear stress $\tau_f$ at the river bed is represented as
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q_w}{w_{cf}} \Big)^{\frac{2}{3}} \,\frac{\partial z_c}{\partial x}^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\frac{\partial z_c}{\partial x}$ is the channel slope, and $f_p$ is the friction factor for subaerial systems.
Channel width $w_{cf}$ is 
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q_w^e,
\end{equation}
% 
where $k$ is a constant and $e$ is an exponent commonly equal to $\frac{1}{2}$ \citep{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
% 
As in Section~\ref{sect:sub_mode}, the width integrated shear stress is $\tau_{tf}=w_{cf}\,\tau_f$.

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher in the Swiss Alps and the Leverett glacier on the Greenland Ice Sheet.
The hydrographs applied to the subglacial and subaerial parameterizations represent the subglacial flow of water as it leaves the glacier and moves into a subaerial flow in front of the glacier, as water moves through the catchment.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, the Fieschergletscher case is referred to as \alpine{} and is exemplified by relatively thin ice thickness ($h_o$: $225$\,\unit{m}), low discharge ( $\le\,20$\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in discharge.
The Leverett glacier case is referred to as \icesheet{} and is exemplified by thick ice  ($h_o$: $700$\,\unit{m}), high water discharge ($\le\,300$\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in discharge.

The parameterizations are first applied to a reference test case for each glacier that assumes a semi-circular subglacial channel ($\beta=\pi$) and friction factors $f_r$ and $f_p$ are tuned so that the model reproduces reasonable velocities for both  the Greenland Ice Sheet and alpine glaciers \citep[$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$][]{werder2010b,chandler2013}.
Water velocity, shear-stress and width-integrated shear-stress in the subglacial and subaerial systems are presented below.

To characterize the variability in sediment discharge capacity in subglacial compared to subaerial systems further, the parameterizations are applied to range of different channel geometry factors ($\beta$ and $k$) and friction factors $f_r$ and $f_p$.
Model runs with parameter combinations  are culled if their maximum subglacial water velocity exceeds $2\,$\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity exceeds $1\,$\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with some dye tracing measurements \citep[e.g.]{werder2010b,chandler2013}.
The standard deviation of water velocity, shear-stress and width-integrated shear-stress across the season long hydrographs are presented below.
